---
title: "Analysing chromatin experiments with chraw"
author:
- name: Ricardo Rosa Junior
- name: Alejandro Reyes
- name: Sebastian Steinhauser
- name: Christian Kolter
package: chraw
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  chraw analyses chromatin and multi-omic experiments. It extends the
  MultiAssayExperiment class into a ChrawExperiment container that records the
  reference genome and the processing pipeline of an experiment, and it
  collapses each step of a chromatin workflow -- quality control, read
  counting, differential testing, peak annotation and enrichment -- into a
  single function call. This vignette walks through a complete analysis, first
  on ATAC-seq data and then on ChIP-seq data.
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
    %\VignetteIndexEntry{Analysing chromatin experiments with chraw}
    %\VignettePackage{chraw}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Installation

```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("chraw")
```

Annotating peaks additionally requires the annotation packages of the relevant
genome, for instance `BSgenome.Hsapiens.UCSC.hg38`,
`TxDb.Hsapiens.UCSC.hg38.refGene` and `org.Hs.eg.db` for hg38.

# Overview

`chraw` is intended for chromatin data processed with the
[ENCODE pipelines](https://www.encodeproject.org/pipelines/) (ATAC-seq,
CUT&RUN and ChIP-seq), although data produced by other pipelines can also be
handled. The genome assemblies hg19, hg38, mm10 and rn6 are supported.

The central object is a `ChrawExperiment`, an extension of
`r Biocpkg("MultiAssayExperiment")` that additionally records the reference
genome and the pipeline. It is built from a sample sheet listing, for every
sample, the alignment (`bamFile`), the signal track (`bigwigFile`), the peak
calls (`peakFile`) and the pipeline quality-control report (`qcFile`).

```{r libraries}
library(chraw)
library(BiocParallel)
library(ggplot2)
```

Note that all example datasets shipped with the package were reduced to a
single genomic window so that the package stays small; the figures below
therefore show far less signal than a real experiment would.

# ATAC-seq

The ATAC-seq example is taken from
[GSE107023](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE107023):
paired-end samples of breast cancer cells, control versus knockout of the
*TRPS1* gene.

```{r atac-load}
data("ce_atac")

# rewrite_paths() only exists to make the bundled example data findable; it is
# not part of a normal analysis.
ce_atac <- rewrite_paths(ce_atac)

metaDat <- t(data.frame(strsplit(colData(ce_atac)$condition, "_|\\+")))
rownames(metaDat) <- NULL
metaDat <- as.data.frame(metaDat)
colnames(metaDat) <- c("target", "condition")
colData(ce_atac) <- cbind(colData(ce_atac), DataFrame(metaDat))
colData(ce_atac)$sample_alias <- colData(ce_atac)$sampleName
colData(ce_atac)$condition <- c("CTRL", "CTRL", "KO", "KO")
colData(ce_atac)
```

The reference genome and the pipeline are available through accessors:

```{r atac-accessors}
referenceGenome(ce_atac)
pipeline(ce_atac)
```

## Quality control

The quality-control plots read the pipeline's JSON report directly. Every axis
states the quantity and its unit, and the captions record how the numbers were
derived.

```{r atac-qc, fig.width=7, fig.height=5}
plotMappingStats(ce_atac)

plotPCRDuplicateStats(ce_atac)
```

The ATAC-seq pipeline additionally reports the mitochondrial read content and
the fraction of reads falling in annotated regions. Both are specific to that
pipeline: on a ChIP-seq report the two functions below stop with a message
saying which section is missing.

```{r atac-qc2, fig.width=7, fig.height=5}
plotChrMStats(ce_atac)

plotFracReadsInAnnot(ce_atac)
```

The fragment length distribution is computed from the alignments themselves, so
the bam files have to be indexed first.

```{r atac-fraglen, fig.width=7, fig.height=5}
indexBam(ce_atac)

ce_atac <- addFragmentLengthDist(
  ce_atac,
  param = csaw::readParam(pe = "both", restrict = "chr8"),
  BPPARAM = SerialParam())

plotFragmentLengthDist(ce_atac) +
  coord_cartesian(xlim = c(0, 450))
```

## Signal around peaks

```{r atac-peaks}
Samples <- ce_atac$sampleName
Peaks <- importNarrowPeaks(ce_atac, includeSamples = Samples, merge = TRUE)
mergedPeaksLarge <- resize(Peaks, 2000, fix = "center")
mergedPeaksLarge <- mergedPeaksLarge[
  seqnames(mergedPeaksLarge) == "chr8" &
    start(mergedPeaksLarge) >= 115000152 &
    end(mergedPeaksLarge) <= 115999730]
names(Peaks) <- sprintf("Peak%0.5d", seq_along(Peaks))
```

`plotMetaProfileHeatmaps()` puts every sample on a shared colour scale, so the
panels can be compared with each other, and labels the x axis in base pairs
relative to the centre of the regions.

```{r atac-meta, fig.width=8, fig.height=5}
plotMetaProfileHeatmaps(
  ce_atac, includeSamples = Samples,
  mergedPeaksLarge, sampleLabelColumn = "sample_alias", bin.num = 100)
```

## Counting reads

```{r atac-counts}
ce_atac <- addCountExperiment(
  ce_atac, regions = Peaks,
  includeSamples = Samples,
  name = "Peaks", BPPARAM = SerialParam())
experiments(ce_atac)
```

```{r atac-corr, fig.width=7, fig.height=5}
plotSampleCorrelations(
  ce_atac, experimentName = "Peaks",
  annotationColumns = "condition",
  sampleLabelColumn = "sample_alias")
```

```{r atac-pca, fig.width=6, fig.height=6}
plotPCA(ce_atac, experimentName = "Peaks",
        colourGrouping = "condition",
        transformation = asinh)
```

The ATAC-seq example stops here. Its reads cover a single megabase of chr8 in
which control and knockout do not differ, so it has nothing to show for the
differential step; that part of the workflow is demonstrated on the ChIP-seq
data below, where the example does carry a real difference.

# ChIP-seq

The ChIP-seq example is taken from
[GSE104060](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE104060):
paired-end H3K27ac samples, control versus an agonist of *NR1I3*, after 3 and
27 hours.

```{r chip-load}
data("ce_chipseq")
ce_chipseq <- rewrite_paths(ce_chipseq)

metaDat <- t(data.frame(strsplit(colData(ce_chipseq)$condition, "_|\\+")))
rownames(metaDat) <- NULL
metaDat <- as.data.frame(metaDat)
colnames(metaDat) <- c("donor", "target", "condition")
colData(ce_chipseq) <- cbind(colData(ce_chipseq), DataFrame(metaDat))
colData(ce_chipseq)$sample_alias <- colData(ce_chipseq)$sampleName
colData(ce_chipseq)$condition <- c("CTRL_3h", "CTRL_3h",
                                   "CTRL_27h", "CTRL_27h",
                                   "agonist_3h", "agonist_3h",
                                   "agonist_27h", "agonist_27h")
# Exposure time, taken from the condition rather than from a column position.
colData(ce_chipseq)$Time <- sub(".*_", "", colData(ce_chipseq)$condition)
ce_chipseq$sample_alias <- gsub("H3K27ac_NR1I3_", "", ce_chipseq$sample_alias)
colData(ce_chipseq)
```

## Quality control

```{r chip-qc, fig.width=8, fig.height=5}
plotPCRDuplicateStats(ce_chipseq)

plotMappingStats(ce_chipseq)
```

## Signal around peaks

```{r chip-meta, fig.width=8, fig.height=5}
# One replicate per condition
Samples <- rownames(colData(ce_chipseq))[c(1, 3, 5, 7)]

Peaks <- importNarrowPeaks(ce_chipseq, includeSamples = Samples, merge = TRUE)
mergedPeaksLarge <- resize(Peaks, 2000, fix = "center")
mergedPeaksLarge <- mergedPeaksLarge[
  seqnames(mergedPeaksLarge) == "chr6" &
    start(mergedPeaksLarge) >= 67090000 &
    end(mergedPeaksLarge) <= 67170000]
names(Peaks) <- sprintf("Peak%0.5d", seq_along(Peaks))

plotMetaProfileHeatmaps(
  ce_chipseq, includeSamples = Samples,
  mergedPeaksLarge, sampleLabelColumn = "sample_alias", bin.num = 100)
```

## Counting reads

```{r chip-counts}
Samples <- ce_chipseq$sampleName
ce_chipseq <- addCountExperiment(
  ce_chipseq, regions = Peaks,
  includeSamples = Samples,
  name = "Peaks", BPPARAM = SerialParam())
experiments(ce_chipseq)
```

```{r chip-corr, fig.width=9.5, fig.height=5.5}
colors <- list(
  "condition" = c("agonist_3h" = "#881bce", "CTRL_3h" = "#24bda5",
                  "agonist_27h" = "#0c1c58", "CTRL_27h" = "#1134b4"),
  "Time" = c("3h" = "#c98a16", "27h" = "#94650f"))

plotSampleCorrelations(
  ce_chipseq, experimentName = "Peaks",
  annotationColumns = c("condition", "Time"),
  annotationColors = colors,
  sampleLabelColumn = "sample_alias")
```

```{r chip-pca, fig.width=6, fig.height=6}
plotPCA(ce_chipseq, experimentName = "Peaks",
        colourGrouping = "condition",
        transformation = asinh)
```

## Coverage tracks

```{r chip-tracks, fig.width=7, fig.height=3}
plottingRegion <- GRanges("chr6:67090000-67170000")
covData <- importAndAverage(object = ce_chipseq,
                            includeSamples = Samples,
                            plottingRegion = plottingRegion)
head(covData, n = 5)

covData$replicate_group <- gsub("H3K27ac_NR1I3_", "", covData$replicate_group)

plotCovFromDF(covData = covData, replicate_group = "agonist_27h",
              plottingRegion = plottingRegion, cl = "black")
```

## Choosing the regions to test

H3K27ac is a broad mark, and the MACS2 peaks of this example have a median
width of about 270 bp. Splitting a broad domain into many narrow peaks divides
its reads among low-count regions, and the differential test then has very
little to work with. Counting the same reads in wider windows keeps the depth
together.

`getGenomicBins()` partitions a whole reference genome into windows of a given
size. Here the analysis is restricted to the locus the example data covers, so
the windows are built directly on that interval — deliberately wider than the
region the reads span, so that the windows also reach nearby genes and can be
annotated later.

```{r chip-bins}
bins <- unlist(tile(GRanges("chr6", IRanges(66500000, 67700000)), width = 5000))
names(bins) <- sprintf("bin%0.5d", seq_along(bins))

ce_chipseq <- addCountExperiment(ce_chipseq, regions = bins, name = "Bins",
                                 BPPARAM = SerialParam())
experiments(ce_chipseq)
```

## Differential testing

Both the agonist versus control comparisons, at 3 and at 27 hours:

```{r chip-diff, fig.width=8, fig.height=5}
ce_chipseq <- testForDiffSignal(
  ce_chipseq, experimentName = "Bins",
  design = ~condition,
  contrasts = list(agonist_27h = c("condition", "agonist_27h", "CTRL_27h"),
                   agonist_3h = c("condition", "agonist_3h", "CTRL_3h")))

getDiffResultSummary(ce_chipseq)

plotMA(ce_chipseq, experimentName = "Bins")

plotVolcano(ce_chipseq, experimentName = "Bins")
```

The windows that reach significance can be pulled out as a `GRanges`:

```{r chip-pull}
diffRes27h <- pullDiffResults(ce_chipseq, experimentName = "Bins",
                              contrastName = "agonist_27h")

# The result columns are prefixed with the contrast name.
padj27h <- diffRes27h$agonist_27h_padj
significant <- diffRes27h[!is.na(padj27h) & padj27h < 0.1]
significant[order(significant$agonist_27h_padj)]
```

Two contrasts can also be compared against each other. The axis titles name the
contrast and the experiment they came from, and the colour legend follows the
`colorByFDR` argument.

```{r chip-scatter, fig.width=6, fig.height=5}
plotDiffScatter(object = ce_chipseq,
                contrasts = c("agonist_3h", "agonist_27h"),
                experimentNames = c("Bins", "Bins"))
```

## Annotating the regions

`annotateExperimentRegions()` needs the annotation packages of the assembly.
By default it does not download the ENCODE ChromHMM segmentation; set
`download = TRUE` to fetch it (the file is then cached with
`r Biocpkg("BiocFileCache")`).

```{r chip-annot, eval=requireNamespace("TxDb.Mmusculus.UCSC.mm10.knownGene", quietly=TRUE) && requireNamespace("org.Mm.eg.db", quietly=TRUE) && requireNamespace("BSgenome.Mmusculus.UCSC.mm10", quietly=TRUE)}
ce_chipseq <- annotateExperimentRegions(ce_chipseq, "Bins", download = FALSE)

enrichRes <- enrichAnno(ce_chipseq, "Bins",
                        "agonist_27h", "simple_annotation",
                        padjThres = 0.1, log2FcThrs = 0, foreground = "up",
                        background = "all")

if (!is.null(enrichRes)) {
  plotEnrichResults(enrichRes)
} else {
  message("No annotation category was enriched in the example data.")
}
```

# Session information

```{r session}
sessionInfo()
```
